---
title: "scRNA-seq Fibrosis Dataset Analysis"
format: html
editor: visual
---

### Abstract

Currently there are no effective antifibrotic therapies for liver cirrhosis, a major killer worldwide. To obtain a cellular resolution of directly-relevant pathogenesis and to inform therapeutic design, we profile the transcriptomes of over 100,000 human single cells, yielding molecular definitions for non-parenchymal cell types present in healthy and cirrhotic human liver. We uncover a novel scar-associated TREM2^+^CD9^+^ macrophage subpopulation, which expands in liver fibrosis, differentiates from circulating monocytes and is pro-fibrogenic. We also define novel ACKR1^+^ and PLVAP^+^ endothelial cells which expand in cirrhosis, are topographically scar-restricted and enhance leucocyte transmigration. Multi-lineage ligand-receptor modelling of interactions between the novel scar-associated macrophages, endothelial cells and PDGFRα^+^ collagen-producing mesenchymal cells reveals intra-scar activity of several pro-fibrogenic pathways including TNFRSF12A, PDGFR and NOTCH signalling. Our work dissects unanticipated aspects of the cellular and molecular basis of human organ fibrosis at a single-cell level, and provides the conceptual framework required to discover rational therapeutic targets in liver cirrhosis.

### Dataset Design

Single-cell transcriptomic data were analyzed from **5 healthy livers, 5 cirrhotic livers, and 4 PBMC samples** from cirrhotic patients. In parallel, single-cell transcriptomic data were also evaluated on **hepatic macrophages isolated from healthy mice and from mice following chronic carbon tetrachloride administration** to model the disease.

### Analysis Design

The primary objective of this study was to isolate reliable target genes in the human cirrhotic liver using scRNA-seq data. **1) Data retrieval from the provided accession number GSE136103 and creation of Seurat object** was first performed to establish the computational framework. **2) QC of the scRNA-seq using the "Seurat" package** was then conducted to filter out low-quality cells, followed immediately by **3) Batch Correction and Integration** to eliminate technical variation. **4) Dimension Reduction** was executed via PCA and UMAP/t-SNE for 2D mapping, enabling the **5) Visualization of Markers associated with the data and annotation of celltypes**. **6) Assess which of celltypes are more profound in cirrhotic tissue vs healthy tissue**.

To understand transcriptomic shifts, the pipeline was designed to **7) Conduct Gene Differentials for each celltype and overall for all cells**, followed by **8) Finding Gene Ontology Terms and Building GO terms plots** to resolve functional biological insights. Given their critical roles in liver pathology, specialized sub-analyses were performed, including **9) Endothelial subcluster annotation and analysis**, **10) Mesenchymal subcluster annotation and analysis**, and **11) Macrophage subcluster annotation and analysis**. A **12) Random forest implementation to select for features that classify cirrhosis cells and healthy cells accurately on Endothelial, Mesenchyme and MP cluster** was then utilized to leverage machine learning for biomarker discovery. This culminated in the **13) Identificaton of features that are improtant and up-regulated via DEGs** to yield a refined, high-confidence set of prioritized targets for liver cirrhosis.

### Load Necessary R Packages for this analysis

```{r, include=F}
library(Seurat)
library(harmony)
library(DoubletFinder)
library(tidyverse)
library(tidymodels)
library(scCustomize) # for plotting
library(patchwork)
library("org.Hs.eg.db")
library(clusterProfiler)
library(rsample)
library(recipes)
library(parsnip)
library(workflows)
library(yardstick)
set.seed(2702)
```

### Set Working Directory

Set the directory to where the scRNA-seq data is located. Please make sure no other files exist in this directory as it will not be able to create seurat objects.

```{r, include=F}
setwd("/projects/standard/lniedern/ali00724/Liver_Fibrosis_Data")

```

### Part 1: Data Curation

For each sample a Seurat object has been created.

```{r, include = F}

# 1. List all raw data files in the specified directory and set the working directory
all_files <- list.files(path = "/projects/standard/lniedern/ali00724/Liver_Fibrosis_Data")
setwd("/projects/standard/lniedern/ali00724/Liver_Fibrosis_Data")

# 2. Extract unique sample prefixes by stripping out the standard 10X file extensions
prefixes <- unique(sub("_(barcodes\\.tsv|genes\\.tsv|matrix\\.mtx)$", "", all_files))

# Initialize an empty list to store individual Seurat objects
seurat_list <- list()

# 3. Loop through each unique prefix to load data and construct Seurat objects
for (prefix in prefixes) {
  message("Currently processing: ", prefix)
  
  # Construct specific file paths for the MTX matrix, features (genes), and barcodes
  matrix_file   <- paste0(prefix, "_matrix.mtx")
  features_file <- paste0(prefix, "_genes.tsv")
  barcodes_file <- paste0(prefix, "_barcodes.tsv")
  
  # Read the sparse matrix components using Seurat's ReadMtx function
  counts <- ReadMtx(mtx = matrix_file, features = features_file, cells = barcodes_file)
  
  # Create the Seurat object, filtering out genes expressed in fewer than 3 cells
  obj <- CreateSeuratObject(counts = counts, project = prefix, min.cells = 3)

  # Assign the experimental condition (Healthy vs Cirrhotic) based on the file prefix naming convention
  if (grepl("healthy", prefix)) {
    obj$Condition <- "Healthy"
  } else {
    obj$Condition <- "Cirrhotic"
  }
  
  # Store the unique sample ID within the object metadata
  obj$SampleID <- prefix
  
  # Append the fully annotated Seurat object to the master list
  seurat_list[[prefix]] <- obj
}
```

#### Calculate Mitochondrial Percentage for each cell in each sample

```{r}
seurat_list <- lapply(X = seurat_list, FUN = function(x) {
  x <- PercentageFeatureSet(x, pattern = "^MT-", col.name = "percent.mt")
})
```

#### Visualize the number of RNA counts

To evaluate sample quality, the RNA counts will be visualized using a violin plot to establish an upper threshold for filtering out anomalies. This step serves as the primary filter to eliminate potential outliers, doublets, or triplets from the dataset.

For this phase of the analysis, only a single representative sample will be inspected, as this specific sample will serve as the anchor reference for downstream multi-sample integration. Other standard quality metrics will not be viewed or adjusted during this visual inspection, as the filtering thresholds have already been explicitly defined by the original publication associated with this dataset.

```{r}
VlnPlot(seurat_list[[1]], features = c("nFeature_RNA")) 
```

### Part 2: Sample Quality Control/Pre-processing

To refine the dataset, cells with fewer than 300 detected genes (`nFeature_RNA`) or a mitochondrial gene percentage exceeding 30% will be removed to eliminate low-quality or dying cells. Conversely, an upper threshold will be applied to exclude cells with more than 4,000 detected genes, targeting potential multiplets.

Following these manual thresholds, DoubletFinder will be implemented to computationally identify and remove remaining doublets. Based on community benchmarks and empirical data discussed on relevant GitHub repositories for this technology, an expected doublet rate of 3% will be utilized based on the typical cell density per sample in this dataset.

```{r, include=F}

# Process each Seurat object in the list sequentially using lapply
seurat_list <- lapply(X = seurat_list, FUN = function(x) {
  
  # 1. Filter out low-quality cells, dying cells, and potential multiplets based on predefined thresholds
  x <- subset(x, nFeature_RNA > 300 & percent.mt < 30 & nFeature_RNA < 4000)
  
  # 2. Normalize gene expression counts using global scaling normalization
  x <- NormalizeData(x)
  
  # 3. Identify the top 2000 highly variable genes that drive biological variation across cells
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
  
  # 4. Scale and center the expression matrix so that genes have equal weight in downstream analysis
  x <- ScaleData(x)
  
  # 5. Perform Principal Component Analysis (PCA) for linear dimensionality reduction
  x <- RunPCA(x)
  
  # 6. Run Uniform Manifold Approximation and Projection (UMAP) using the first 10 PCs for 2D visualization
  x <- RunUMAP(x, dims = 1:10)
  
  # 7. Calculate the expected number of doublets per sample assuming a 3% doublet rate
  nExp_poi <- round(0.03 * nrow(x@meta.data))
  
  # 8. Run DoubletFinder to computationally predict doublets using standard hyperparameter settings
  x <- doubletFinder(x, PCs = 1:10, pN = 0.25, pK = 0.09, nExp = nExp_poi, reuse.pANN = NULL, sct = FALSE)
  
  # 9. Standardize the metadata column name where DoubletFinder stores classification outputs
  colnames(x@meta.data)[8] = "Doublet_info"
  
  # 10. Subset the Seurat object to retain only high-confidence single cells, discarding predicted doublets
  x = subset(x, Doublet_info == "Singlet")
})

```

### Part 3: Batch Correction & Integration of Samples

Seurat integration steps do batch correction thus we do not need to take any steps. I used the first sample as reference to do the integration.

```{r, include=F}
# 1. Select the top 2000 highly variable features that are consistently shared across all samples
features <- SelectIntegrationFeatures(object.list = seurat_list, nfeatures = 2000)

# 2. Identify integration anchors across datasets, utilizing the first sample as the reference anchor
seurat.anchors <- FindIntegrationAnchors(object.list = seurat_list, anchor.features = features, reference = 1)

# 3. Mathematically batch-correct and merge the samples based on the identified anchor points
seurat.combined <- IntegrateData(anchorset = seurat.anchors)

# 4. Set the default assay to "integrated" to ensure downstream clustering and dimensionality reduction utilize the batch-corrected values
seurat.combined@active.assay <- "integrated"
```

### Part 4: Dimensional Reduction, Clustering

We have scaled the batch corrected counts, run principal components analysis and used harmony which is notorious for robust clustering. We calculated neighbors based on 15 harmony dimensions and clusters were calculated at 0.4. Finally, a UMAP was created based on the harmony dimensions.

```{r, include  = F}
# 1. Scale and center the integrated expression data across all cells
seurat.combined = ScaleData(seurat.combined)

# 2. Perform initial linear dimensionality reduction using Principal Component Analysis (PCA)
seurat.combined = RunPCA(seurat.combined)

# 3. Apply Harmony integration to further blend samples and remove batch effects across experimental runs (orig.ident)
seurat.combined = RunHarmony(seurat.combined, group.by.vars = "orig.ident")

# 4. Construct a Shared Nearest Neighbor (SNN) graph based on the first 15 Harmony dimensions
seurat.combined = FindNeighbors(seurat.combined, dims = 1:15, reduction = "harmony")

# 5. Cluster cells using the Louvain algorithm at a moderate resolution of 0.4
seurat.combined = FindClusters(seurat.combined, resolution = 0.4)

# 6. Run non-linear dimensionality reduction (UMAP) on the Harmony embeddings for 2D visualization
seurat.combined = RunUMAP(seurat.combined, dims = 1:15, reduction = "harmony")

# 7. Generate a UMAP dimensional reduction plot with cluster labels displayed
DimPlot(seurat.combined, label = T)

# 8. Revert the active assay to "RNA" to ensure downstream differential expression uses raw/normalized counts
seurat.combined@active.assay = "RNA"

# 9. Merge separated data layers (e.g., across batches) into a unified matrix for comprehensive analysis
seurat.combined = JoinLayers(seurat.combined)

```

### Part 5: Celltype annotation

Cells were annotated based on the markers provided by article associated with this dataset.

#### Curate Marker Vector for each celltype

```{r, include = F}
mp_markers      <- c("FCN1", "VCAN", "C1QC", "IL1B", "MNDA", "CPVL", "MS4A7", "MSR1", "CD163", "CYBB", "CSF2RA", "CLEC10A")
pdc_markers     <- c("IL3RA", "LRRC26", "PTCRA", "CLEC4C")
ilc_markers     <- c("KLRF1", "KLRC1", "PRF1", "FGFBP2")
cycling_markers <- c("MKI67", "TOP2A", "UBE2C", "RRM2")
t_markers       <- c("TRAC", "CD2")
b_markers       <- c("CD79A", "MS4A1")
plasma_markers  <- c("FCRL5", "JSRP1")
mast_markers    <- c("TPSAB1", "CPA3")
endo_markers    <- c("CLEC14A", "PLVAP", "EGFL7", "VWF", "HSPG2", "EMCN", "FCN3", "OIT3", "CLEC4G", "CLEC4M")
epithelia_markers    <- c("ALB", "F2", "KNG1", "PRAP1", "TTR", "HSD17B6", "ORM1", "ITIH1", "TF", "CYP4A11", "APOC2", "AGXT")
mesenchyma_markers    <- c("TAGLN", "RGS5", "ACTA2", "PDGFRB", "TIMP1", "LUM", "COL3A1", "PDGFRA", "COL1A2", "CCL19")
```

#### Visualize the Mononuclear Phagocyte Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = mp_markers) + RotatedAxis() # Cluster 3,6,8
```

#### Visualize the Dendritic Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = pdc_markers) + RotatedAxis()
# Cluster 16 pDC
```

#### Visualize the Innate Lymphoid Cell Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = ilc_markers) + RotatedAxis()
# Cluster 5,1 ILC
```

#### Visualize the Cyclic Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = cycling_markers) + RotatedAxis()
```

#### Visualize the T Cell Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = t_markers) + RotatedAxis()
# Cluster 0,2 T Cell
```

#### Visualize the B Cell Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = b_markers) + RotatedAxis()
# Cluster 10 B Cell

```

#### Visualize the Plasma Cell Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = plasma_markers) + RotatedAxis()
# Cluster 12 Plasma Cell
```

#### Visualize the Mast Cell Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = mast_markers) + RotatedAxis()
# Cluster 13 Mast Cell

```

#### Visualize the Endothelial Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = endo_markers) + RotatedAxis()
# Cluster 4,7 Endothelial 

```

#### Visualize the Epithelial Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = epithelia_markers) + RotatedAxis()
# Cluster 15 Epithelial

```

#### Visualize the Mesenchymal Marker in a Dot Plot

```{r}
DotPlot(seurat.combined, assay = "RNA", features = mesenchyma_markers) + RotatedAxis()
# Cluster 13,11 Mesenchymal

```

#### Annotate the Celltypes

```{r}
celltype = c("T Cell", "ILC", "T Cell", "MP", "Endothelial", "ILC", "MP", "Endothelial", "MP", "Epithelial", "B Cell", "Mesenchyme", "Plasma Cell", "Mast Cell", "Cycling", "Epithelial", "pDC")
names(celltype) <- levels(seurat.combined)
seurat.combined <- RenameIdents(seurat.combined, celltype)


```

#### Visualize the Dimensional Plot with Celltype Annotations

```{r}
DimPlot(seurat.combined, label = T, split.by = "Condition")
DimPlot(seurat.combined, label = T)
```

### Part 6: Assess cell composition by Condition

Based on the cellular composition analysis, cirrhotic tissue exhibits a marked increase in **Endothelial cells, Mononuclear Phagocytes (MPs), Epithelial cells, B Cells, Plasma Cells, Mast Cells, Cycling Cells, and Dendritic Cells** compared to healthy tissue. Conversely, healthy tissue retains a higher proportion of **T Cells, Innate Lymphoid Cells (ILCs), and Mesenchymal Cells**.

```{r}
# Generate frequency table with meaningful column names
cell_counts <- table(seurat.combined@active.ident, seurat.combined$Condition)
cell_counts_df <- as.data.frame(cell_counts)

# Rename default 'Var1' and 'Var2' columns for clarity
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var1"] <- "CellType"
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var2"] <- "Condition"

# Initialize and assign total cell counts per condition
cell_counts_df$TotalCells = 1
cell_counts_df[cell_counts_df$Condition == "Cirrhotic", ]$TotalCells = 24716
cell_counts_df[cell_counts_df$Condition == "Healthy", ]$TotalCells = 33848

# Calculate percentage
cell_counts_df$Percentage = cell_counts_df$Freq / cell_counts_df$TotalCells * 100

# Plot using the updated, clean variable names
ggplot(cell_counts_df, aes(y = Percentage, x = CellType, fill = Condition)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  RotatedAxis() + 
  xlab("Cell Type") + 
  ylab("Percentage of Cells") + 
  guides(fill = guide_legend(title = "Condition"))



```

### Part 7: Conduct Gene Differential on all cells by condition ( Cirrhotic vs Health)

```{r}
# Find markers comparing Cirrhotic vs Healthy
cirrhotic_vs_healthy_markers = FindMarkers(seurat.combined, assay = "RNA", group.by = "Condition", ident.1 = "Cirrhotic", ident.2 = "Healthy")

# Filter for significantly up-regulated genes
cirrhotic_up_regulated = cirrhotic_vs_healthy_markers %>% filter(avg_log2FC > 0 & p_val_adj < 0.05)

# Check the dimensions (4413 genes identified as upregulated in cirrhotic tissue)
dim(cirrhotic_up_regulated)
```

### Part 8: Finding Gene Ontology Terms for Each Celltype Based on Significant DEGs for each celltype (Cirrhotic vs Healthy)

```{r, include=F}
# set identities if needed
# Idents(seurat.combined) <- "celltype"


organism <- org.Hs.eg.db

# all idents
idents <- levels(Idents(seurat.combined))

# output lists
marker_list <- list()
gsea_list <- list()

for (id in idents) {
  
  cat("\nProcessing:", id, "\n")
  
  # subset current ident
  sub_obj <- subset(seurat.combined, idents = id)
  
  # ensure both conditions exist
  conds <- unique(sub_obj$Condition)
  
  if (!all(c("Cirrhotic", "Healthy") %in% conds)) {
    cat("Skipping", id, "- missing one condition\n")
    next
  }
  
  # differential expression
  mks <- FindMarkers(
    sub_obj,
    assay = "RNA",
    group.by = "Condition",
    ident.1 = "Cirrhotic",
    ident.2 = "Healthy",
    logfc.threshold = 0,
    min.pct = 0
  )
  
  # save all markers
  marker_list[[id]] <- mks
  
  # keep significant genes only
  df <- mks %>%
    rownames_to_column("gene") %>%
    filter(!is.na(avg_log2FC)) %>%
    filter(p_val_adj < 0.05) %>%
    dplyr::select(gene, avg_log2FC)
  
  # skip if too few significant genes
  if (nrow(df) < 10) {
    cat("Skipping GSEA for", id, "- too few significant genes\n")
    next
  }
  
  # create ranked vector
  original_gene_list <- df$avg_log2FC
  names(original_gene_list) <- df$gene
  
  # remove NA
  gene_list <- na.omit(original_gene_list)
  
  # remove duplicate genes
  gene_list <- gene_list[!duplicated(names(gene_list))]
  
  # sort decreasing (required for GSEA)
  gene_list <- sort(gene_list, decreasing = TRUE)
  
  cat("Running GSEA for", id, "with", length(gene_list), "genes\n")
  
  # GSEA GO
  gse <- gseGO(
    geneList       = gene_list,
    ont            = "ALL",
    keyType        = "SYMBOL",
    nPerm          = 10000,
    minGSSize      = 3,
    maxGSSize      = 800,
    pvalueCutoff   = 0.05,
    verbose        = TRUE,
    OrgDb          = organism,
    pAdjustMethod  = "none"
  )
  
  # save GSEA result
  gsea_list[[id]] <- gse
}

```

#### Top 10 Go Terms Plot for Each Celltype

```{r, include=T}

# Loop through each cell type in the list by its name
for (cell_type in names(gsea_list)) {
  
  # Create a clean file name (e.g., "GSEA_T_Cell_dotplot.pdf")
  file_name <- paste0("GSEA_", gsub(" ", "_", cell_type), "_dotplot.pdf")
  
  # 1. Open the PDF graphics device with height 8 and width 8
  pdf(file_name, height = 8, width = 8)
  
  # 2. Generate and print the dotplot for the current cell type
  print(
    dotplot(gsea_list[[cell_type]], 
            showCategory = 10, 
            title = paste("Enriched Pathways", cell_type), 
            split = ".sign") + 
    facet_grid(.~.sign)
  )
  
  # 3. Close the device to save the file
  dev.off()
  
  message("Successfully saved: ", file_name)
}
```

### Part 9: Endothelial Celltype Subclustering

```{r, include = F}
# 1. Isolate the main Endothelial cell cluster from the integrated dataset
Endothelial <- subset(seurat.combined, idents = "Endothelial")

# 2. Reset the active assay to "RNA" for raw/normalized expression metrics
DefaultAssay(Endothelial) <- "RNA"

# 3. Identify the top 2000 highly variable features specific to the endothelial lineage
Endothelial <- FindVariableFeatures(Endothelial, selection.method = "vst", nfeatures = 2000)

# 4. Scale and center the expression matrix for downstream dimensionality reduction
Endothelial <- ScaleData(Endothelial)

# 5. Compute the first 30 Principal Components (PCs) using the lineage-specific variable features
Endothelial <- RunPCA(Endothelial, features = VariableFeatures(object = Endothelial), npcs = 30)

# 6. Construct an SNN graph using the top 15 principal components
Endothelial <- FindNeighbors(Endothelial, dims = 1:15)

# 7. Cluster the endothelial cells using a conservative resolution of 0.2 to identify major subclusters
Endothelial <- FindClusters(Endothelial, resolution = 0.2)

# 8. Generate a UMAP non-linear dimensional reduction layout for 2D visualization
Endothelial <- RunUMAP(Endothelial, dims = 1:15)

# 9. Generate an alternative t-SNE dimensional reduction layout for comprehensive cluster viewing
Endothelial <- RunTSNE(Endothelial, dims = 1:15)
```

#### Endothelial Subcluster Marker Dotplot and Annotation

```{r}
# 1. Define a comprehensive panel of canonical endothelial lineage and subset markers
all_endo_markers <- c(
  "CLEC4G", "CLEC4M", "STAB2", "MRC1", "CD14", "LYVE1", "CCL21", "PROX1", "PDPN", 
  "CPE", "CD320", "BMPER", "RSPO3", "WNT2", "LHX6", "HTR2B", "AIF1L", "SOX17", 
  "PLVAP", "VWA1", "ACKR1"
)

# 2. Visualize marker expression profiles across clusters via a dot plot, rotating x-axis labels for readability
DotPlot(Endothelial, assay = "RNA", features = all_endo_markers) + RotatedAxis()

# 3. Create a vector mapping the numerical clusters to specific endothelial subclusters
celltype <- c("Endo6", "Endo5", "Endo6", "Endo7", "Endo1", "Endo3", "Endo2", "Endo4", "Endo6")

# 4. Assign the original cluster numbers as names to the mapping vector
names(celltype) <- levels(Endothelial)

# 5. Rename the active identities in the Endothelial Seurat object to reflect the new subcluster labels
Endothelial <- RenameIdents(Endothelial, celltype)
```

#### Endothelial Subcluster and Cell Composition Visualization

Endo6, Endo7 and Endo2 are subcluster that are more prominent in Cirrhotic tissue in comparison to healthy tissue.

```{r}
DimPlot(Endothelial, label = TRUE, split.by = "Condition")


# Generate frequency table with meaningful column names
cell_counts <- table(Endothelial@active.ident, Endothelial$Condition)
cell_counts_df <- as.data.frame(cell_counts)

# Rename default 'Var1' and 'Var2' columns for clarity
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var1"] <- "CellType"
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var2"] <- "Condition"

# Initialize and assign total cell counts per condition
cell_counts_df$TotalCells = 1
cell_counts_df[cell_counts_df$Condition == "Cirrhotic", ]$TotalCells = 4409
cell_counts_df[cell_counts_df$Condition == "Healthy", ]$TotalCells = 2580

# Calculate percentage
cell_counts_df$Percentage = cell_counts_df$Freq / cell_counts_df$TotalCells * 100

# Plot using the updated, clean variable names
ggplot(cell_counts_df, aes(y = Percentage, x = CellType, fill = Condition)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  RotatedAxis() + 
  xlab("Cell Type") + 
  ylab("Percentage of Cells") + 
  guides(fill = guide_legend(title = "Condition"))

```

### Part 10: Mesenchymal Subclustering

```{r, include=F}
# 1. Isolate the main Mesenchymal cell cluster from the integrated dataset
Mesenchymal <- subset(seurat.combined, idents = "Mesenchyme")

# 2. Reset the active assay to "RNA" for raw/normalized expression metrics
DefaultAssay(Mesenchymal) <- "RNA"

# 3. Identify the top 2000 highly variable features specific to the mesenchymal lineage
Mesenchymal <- FindVariableFeatures(Mesenchymal, selection.method = "vst", nfeatures = 2000)

# 4. Scale and center the expression matrix for downstream dimensionality reduction
Mesenchymal <- ScaleData(Mesenchymal)

# 5. Compute the first 30 Principal Components (PCs) using the lineage-specific variable features
Mesenchymal <- RunPCA(Mesenchymal, features = VariableFeatures(object = Mesenchymal), npcs = 30)

# 6. Construct an SNN graph using the top 15 principal components
Mesenchymal <- FindNeighbors(Mesenchymal, dims = 1:15)

# 7. Cluster the mesenchymal cells using a resolution of 0.2 to identify major subclusters
Mesenchymal <- FindClusters(Mesenchymal, resolution = 0.2)

# 8. Generate a UMAP non-linear dimensional reduction layout for 2D visualization
Mesenchymal <- RunUMAP(Mesenchymal, dims = 1:15)

# 9. Generate an alternative t-SNE dimensional reduction layout for comprehensive cluster viewing
Mesenchymal <- RunTSNE(Mesenchymal, dims = 1:15)
```

#### Mesenchymal Marker Dotplot

```{r}
paper_meso_markers <- c(
  "MYH11", "RGS5",                       # VSMC
  "PDGFRA",                              # HSC
  "COL1A1", "COL3A1", "TIMP1", "CCL2",   # SAMes (Scar-Associated)
  "KRT19", "PDPN", "WT1", "MSLN"         # Mesothelia
)
DotPlot(Mesenchymal, assay = "RNA", features = paper_meso_markers) + RotatedAxis()
```

#### Mesenchymal Subcluster Annotation and Cell Composition

```{r}

celltype <- c("HSC", "VSMC", "SAMes", "Mesothelial")
names(celltype) <- levels(Mesenchymal)
Mesenchymal <- RenameIdents(Mesenchymal, celltype)

# 11. Final visualization splits
DimPlot(Mesenchymal, label = TRUE, split.by = "Condition")

```

#### Mesenchymal Subcluster Composition Analysis

SAMes is the subcluster in the Mesenchymal cluster that has more cells in cirrhotic tissue than healthy tissue.

```{r}
# Generate frequency table with meaningful column names
cell_counts <- table(Mesenchymal@active.ident, Mesenchymal$Condition)
cell_counts_df <- as.data.frame(cell_counts)

# Rename default 'Var1' and 'Var2' columns for clarity
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var1"] <- "CellType"
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var2"] <- "Condition"

# Initialize and assign total cell counts per condition
cell_counts_df$TotalCells = 1
cell_counts_df[cell_counts_df$Condition == "Cirrhotic", ]$TotalCells = 344
cell_counts_df[cell_counts_df$Condition == "Healthy", ]$TotalCells = 1290

# Calculate percentage
cell_counts_df$Percentage = cell_counts_df$Freq / cell_counts_df$TotalCells * 100

# Plot using the updated, clean variable names
ggplot(cell_counts_df, aes(y = Percentage, x = CellType, fill = Condition)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  RotatedAxis() + 
  xlab("Cell Type") + 
  ylab("Percentage of Cells") + 
  guides(fill = guide_legend(title = "Condition"))
```

### Part 11: Macrophage Cells Subclustering

```{r, include=F}
# 1. Isolate the Mononuclear Phagocyte (MP) cell cluster from the integrated dataset
MP <- subset(seurat.combined, idents = "MP")

# 2. Reset the active assay to "RNA" for raw/normalized expression metrics
DefaultAssay(MP) <- "RNA"

# 3. Identify the top 2000 highly variable features specific to the MP lineage
MP <- FindVariableFeatures(MP, selection.method = "vst", nfeatures = 2000)

# 4. Scale and center the expression matrix for downstream dimensionality reduction
MP <- ScaleData(MP)

# 5. Compute the first 30 Principal Components (PCs) using the lineage-specific variable features
MP <- RunPCA(MP, features = VariableFeatures(object = MP), npcs = 30)

# 6. Construct an SNN graph using the top 15 principal components
MP <- FindNeighbors(MP, dims = 1:15)

# 7. Cluster the MP cells using a conservative resolution of 0.2 to identify major subclusters
MP <- FindClusters(MP, resolution = 0.2)

# 8. Generate a UMAP non-linear dimensional reduction layout for 2D visualization
MP <- RunUMAP(MP, dims = 1:15)

# 9. Generate an alternative t-SNE dimensional reduction layout for comprehensive cluster viewing
MP <- RunTSNE(MP, dims = 1:15)
```

#### MP subcluster Marker Dotplot

```{r}

# 9. Marker Expression DotPlot (Updated with Mononuclear Phagocyte / Macrophage lineage markers)
paper_mp_markers <- c(
  "S100A12", "FCN1", "MNDA", "NBEAL1", "IFITM2", "APOBEC3A", "RGS2", # TMo
  "TREM2", "CD9",                                                   # SAMΦ
  "C1QC", "CD5L", "CD163", "VCAM1", "MARCO", "SIGLEC1", "MERTK",    # KC
  "ARL4C", "FCER1A", "CD1E", "LGALS2"                               # cDC
)
DotPlot(MP, assay = "RNA", features = paper_mp_markers) + RotatedAxis()

```

### Annotation of MP subclusters

```{r}
celltype <- c("Tissue Monocyte", "SAMΦ", "Kupffer cell", "Conventional dendritic cell", "Tissue Monocyte", "Tissue Monocyte", "Tissue Monocyte","Conventional dendritic cell","Tissue Monocyte")
names(celltype) <- levels(MP)
MP <- RenameIdents(MP, celltype)
DimPlot(MP, label = TRUE, split.by = "Condition")

```

#### MP subcluster cell composition analysis

Tissue Monocyte and SAMΦ have more cells in cirrhotic tissue.

```{r}
# Generate frequency table with meaningful column names
cell_counts <- table(MP@active.ident, MP$Condition)
cell_counts_df <- as.data.frame(cell_counts)

# Rename default 'Var1' and 'Var2' columns for clarity
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var1"] <- "CellType"
colnames(cell_counts_df)[colnames(cell_counts_df) == "Var2"] <- "Condition"

# Initialize and assign total cell counts per condition
cell_counts_df$TotalCells = 1
cell_counts_df[cell_counts_df$Condition == "Cirrhotic", ]$TotalCells = 5048
cell_counts_df[cell_counts_df$Condition == "Healthy", ]$TotalCells = 6178

# Calculate percentage
cell_counts_df$Percentage = cell_counts_df$Freq / cell_counts_df$TotalCells * 100

# Plot using the updated, clean variable names
ggplot(cell_counts_df, aes(y = Percentage, x = CellType, fill = Condition)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  RotatedAxis() + 
  xlab("Cell Type") + 
  ylab("Percentage of Cells") + 
  guides(fill = guide_legend(title = "Condition"))
```

#### Part 12: Random Forest Implementation for Important Feature Selection

#### Endothelial Feature Selection

```{r}

# 1. Data Extraction & Reshaping
DefaultAssay(Endothelial) <- "RNA"
Endothelial <- ScaleData(Endothelial)

# Extract and transpose the scaled expression data
endothelial_expr_mat <- Endothelial@assays$RNA$scale.data
endothelial_scaled_df <- as.data.frame(t(endothelial_expr_mat))

# Print dimensions: rows = cells, columns = 2000 highly variable genes
dim(endothelial_scaled_df)

# 2. Annotate Metadata & Format Target Variable
endothelial_scaled_df$cell_type    <- factor(Endothelial$Condition) # Target variable formatted as factor
endothelial_scaled_df$cell_barcode <- rownames(endothelial_scaled_df)


# 3. Data Splitting & Resampling
set.seed(123) # Recommended for reproducibility
endothelial_split <- initial_split(endothelial_scaled_df, strata = "cell_type")
endothelial_train <- training(endothelial_split)
endothelial_test  <- testing(endothelial_split)

# 10-fold cross-validation
endothelial_folds <- vfold_cv(endothelial_train, v = 10)


# 4. Recipe Configuration (Preprocessing)
endothelial_recipe <- recipe(formula = cell_type ~ ., data = endothelial_train) %>%
  update_role(cell_barcode, new_role = "ID") %>%
  step_zv(all_predictors())


# 5. Model Specification & Workflow Pipeline
endothelial_spec <- rand_forest() %>%
  set_engine("randomForest", importance = TRUE) %>%
  set_mode("classification")

endothelial_workflow <- workflow() %>% 
  add_recipe(endothelial_recipe) %>% 
  add_model(endothelial_spec)


# 6. Model Training
endothelial_fit <- fit(endothelial_workflow, data = endothelial_train)


# 7. Model Evaluation (Confusion Matrix)
predict(endothelial_fit, new_data = endothelial_test) %>%
  bind_cols(endothelial_test %>% select(cell_type)) %>%
  conf_mat(truth = cell_type, estimate = .pred_class)


# 8. Feature Importance Visualization
endothelial_fit %>%
  extract_fit_parsnip() %>%
  vip::vip(geom = "col", num_features = 30) + 
  theme_bw(base_size = 14) +
  labs(title = "Endothelial Random Forest Variable Importance")

```

#### Mesenchymal Feature Selection

```{r}
# 1. Data Extraction & Reshaping
DefaultAssay(Mesenchymal) <- "RNA"
Mesenchymal <- ScaleData(Mesenchymal)

# Safely extract ONLY the 2000 highly variable features to save memory
mesenchymal_hvgs <- VariableFeatures(Mesenchymal)
mesenchymal_data_mat <- t(as.matrix(GetAssayData(Mesenchymal, assay = "RNA", slot = "scale.data")[mesenchymal_hvgs, ]))

# Convert to data frame
mesenchymal_scaled_df <- as.data.frame(mesenchymal_data_mat)
# Now, every row is a cell with 2000 genes/features/predictors. 
dim(mesenchymal_scaled_df)

# 2. Annotate Metadata & Format Target Variable
mesenchymal_scaled_df$cell_type    <- factor(Mesenchymal$Condition) # Target variable formatted as factor
mesenchymal_scaled_df$cell_barcode <- rownames(mesenchymal_scaled_df)



set.seed(123) # Recommended for reproducibility
mesenchymal_split <- initial_split(mesenchymal_scaled_df, strata = "cell_type")
mesenchymal_train <- training(mesenchymal_split)
mesenchymal_test  <- testing(mesenchymal_split)

# 10 fold cross validation
mesenchymal_folds <- vfold_cv(mesenchymal_train, v = 10)


# 4. Recipe Configuration (Preprocessing)
mesenchymal_recipe <- recipe(formula = cell_type ~ ., data = mesenchymal_train) %>%
  update_role(cell_barcode, new_role = "ID") %>%
  step_zv(all_predictors())


# 5. Model Specification & Workflow Pipeline
mesenchymal_spec <- rand_forest() %>%
  set_engine("randomForest", importance = TRUE) %>%
  set_mode("classification")

mesenchymal_workflow <- workflow() %>% 
  add_recipe(mesenchymal_recipe) %>% 
  add_model(mesenchymal_spec)


# 6. Model Training
mesenchymal_fit <- fit(mesenchymal_workflow, data = mesenchymal_train)


# 7. Model Evaluation (Confusion Matrix)
predict(mesenchymal_fit, new_data = mesenchymal_test) %>%
  bind_cols(mesenchymal_test %>% select(cell_type)) %>%
  conf_mat(truth = cell_type, estimate = .pred_class)


# 8. Feature Importance Visualization
mesenchymal_fit %>%
  extract_fit_parsnip() %>%
  vip::vip(geom = "col", num_features = 30) + 
  theme_bw(base_size = 14) +
  labs(title = "Random Forest Variable Importance (Mesenchymal)")
```

#### Mononuclear Phagocyte (MP) Feature Selection

```{r}
# 1. Data Extraction & Reshaping
DefaultAssay(MP) <- "RNA"
MP <- ScaleData(MP)

# Safely extract ONLY the 2000 highly variable features to save memory
mp_hvgs <- VariableFeatures(MP)
mp_data_mat <- t(as.matrix(GetAssayData(MP, assay = "RNA", slot = "scale.data")[mp_hvgs, ]))

# Convert to data frame
mp_scaled_df <- as.data.frame(mp_data_mat)
# Now, every row is a cell with 2000 genes/features/predictors. 
dim(mp_scaled_df)

# 2. Annotate Metadata & Format Target Variable
mp_scaled_df$cell_type    <- factor(MP$Condition) # Target variable formatted as factor
mp_scaled_df$cell_barcode <- rownames(mp_scaled_df)

set.seed(123) # Recommended for reproducibility
mp_split <- initial_split(mp_scaled_df, strata = "cell_type")
mp_train <- training(mp_split)
mp_test  <- testing(mp_split)

# 10 fold cross validation
mp_folds <- vfold_cv(mp_train, v = 10)


# 4. Recipe Configuration (Preprocessing)
mp_recipe <- recipe(formula = cell_type ~ ., data = mp_train) %>%
  update_role(cell_barcode, new_role = "ID") %>%
  step_zv(all_predictors())


# 5. Model Specification & Workflow Pipeline
mp_spec <- rand_forest() %>%
  set_engine("randomForest", importance = TRUE) %>%
  set_mode("classification")

mp_workflow <- workflow() %>% 
  add_recipe(mp_recipe) %>% 
  add_model(mp_spec)


# 6. Model Training
mp_fit <- fit(mp_workflow, data = mp_train)


# 7. Model Evaluation (Confusion Matrix)
predict(mp_fit, new_data = mp_test) %>%
  bind_cols(mp_test %>% select(cell_type)) %>%
  conf_mat(truth = cell_type, estimate = .pred_class)


# 8. Feature Importance Visualization
 mp_fit %>%
  extract_fit_parsnip() %>%
  vip::vip(geom = "col", num_features = 30) + 
  theme_bw(base_size = 14) +
  labs(title = "Random Forest Variable Importance (MP)")
```

### Part 13: Therapeutic Gene Targets

Identification of Gene targets via intersection of important genes identified by random forest and up-regulated gene differential list

#### Endothelial Cirrhotic Target Genes

```{r}

top_20_features <- endothelial_fit %>%
  extract_fit_parsnip() %>%
  vip::vi() %>%             # Returns a tibble with 'Variable' and 'Importance'
  head(30) %>%              # Takes the top 20 highest importance features
  pull(Variable)

Endothelial_DEG_up_regulated = FindMarkers(Endothelial,assay = "RNA", group.by = "Condition", ident.1 = "Cirrhotic", ident.2 = "Healthy") %>% filter(avg_log2FC > 0 & p_val_adj < 0.05)


intersect(rownames(Endothelial_DEG_up_regulated), top_20_features)
```

#### Mesenchymal Cirrhotic Target Genes

```{r}

top_20_features_Mesenchymal <- mesenchymal_fit %>%
  extract_fit_parsnip() %>%
  vip::vi() %>%             # Returns a tibble with 'Variable' and 'Importance'
  head(30) %>%              # Takes the top 20 highest importance features
  pull(Variable)

Mesenchymal_DEG_up_regulated = FindMarkers(Mesenchymal,assay = "RNA", group.by = "Condition", ident.1 = "Cirrhotic", ident.2 = "Healthy") %>% filter(avg_log2FC > 0 & p_val_adj < 0.05)


intersect(rownames(Mesenchymal_DEG_up_regulated), top_20_features_Mesenchymal)

```

#### MP Cirrhotic Target Genes

```{r}

top_20_features_MP <- mp_fit %>%
  extract_fit_parsnip() %>%
  vip::vi() %>%             # Returns a tibble with 'Variable' and 'Importance'
  head(30) %>%              # Takes the top 20 highest importance features
  pull(Variable)

MP_DEG_up_regulated = FindMarkers(MP,assay = "RNA", group.by = "Condition", ident.1 = "Cirrhotic", ident.2 = "Healthy") %>% filter(avg_log2FC > 0 & p_val_adj < 0.05)


intersect(rownames(MP_DEG_up_regulated), top_20_features_MP)

```

|     |
|-----|
|     |
